Shapes analysis for time series.\\ International Conference on Machine Learning (ICML 2024)

### 2024-04-30 NEURIPS24/paper.tex
update paper

%%%%%%%% ICML 2023 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
%\usepackage{tikz}

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2023} with \usepackage[nohyperref]{icml2023} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2023}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \theoremstyle{plain}
% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{proposition}[theorem]{Proposition}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \theoremstyle{definition}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{assumption}[theorem]{Assumption}
% \theoremstyle{remark}
% \newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Shapes analysis for time series.}

\input{header}
\input{def}
\newcommand\sam[1]{\textcolor{blue}{#1}}
\newcommand\thi[1]{\textcolor{orange}{#1}}
\begin{document}

\twocolumn[
\icmltitle{Shapes analysis for time series.\\
           International Conference on Machine Learning (ICML 2024)}%a changer

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2023
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Firstname1 Lastname1}{equal,yyy}
\icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
\icmlauthor{Firstname3 Lastname3}{comp}
\icmlauthor{Firstname4 Lastname4}{sch}
\icmlauthor{Firstname5 Lastname5}{yyy}
\icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
\icmlauthor{Firstname7 Lastname7}{comp}
%\icmlauthor{}{sch}
\icmlauthor{Firstname8 Lastname8}{sch}
\icmlauthor{Firstname8 Lastname8}{yyy,comp}
%\icmlauthor{}{sch}
%\icmlauthor{}{sch}
\end{icmlauthorlist}

\icmlaffiliation{yyy}{Department of XXX, University of YYY, Location, Country}
\icmlaffiliation{comp}{Company Name, Location, Country}
\icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}

\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
%     The most common way to analyze time series is to find a feature representation manually with expert knowledge or automatically with machine learning.
% While unsupervised representation learning on time series proves efficient to reveal cluster structure in a dataset,
%  actual methods lacks precision for analyzing the inter-individual variability whithin a cluster.
Analyzing the inter-individual variability within a cluster of time series is particularly appealing in medicine and biology.
 Individuals have specificities depending on their genes or disease
  while sharing, on average, a similar pattern in the features of their corresponding time series. For instance, an electrocardiogram presents a typical shape of the heartbeat, which varies depending on the individual.
 In this paper, we propose an unsupervised representation method (TS-LDDMM) to analyze inter-individual variability of time series dataset.
 Inspired by the shape analysis literature, we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) to time series, considering the temporal evolution as their distinctive feature.  
 Given a dataset of $N$ individual's time series $(s^j)_{j\in[N]}$,
  we learn a time series of reference $\mathbf{s}_0$ encoding the common part of the individuals as well as diffeomorphisms $(\phi^j)_{j\in[N]}$ encoding the specificities of each individual:
  $s^j$ is seen as the transformation of $\mathbf{s}_0$ by $\phi_j$.
  Then, the parameters $(\alpha_j)_{j\in[N]}$ encoding the diffeomorphisms $(\phi^j)_{j\in[N]}$ are used as the features representation related to each time series $s^j$.
  These features can be post-processed with any statistical or machine learning tools. 
  We demonstrate the advantages of our representation compared to existing methods using synthetic data and real-world examples, motivated by applications in medicine.
  %  After checking the injectivity of the TS-LDDMM representation on synthetic data (potentially to change), we study respiratory shapes of a mouse dataset by applying Principal components analysis to the features $(\alpha_j)_{j\in[N]} $.

  %  Moreover, we show that the method is competitive with supervised methods for classification, as soon as the dataset is homogene enough.
  %   The generalisation of the method to heterogene dataset (several clusters) is relagated for future works.
\end{abstract}

@@

%Idée de parler d'inter individualité, spécialiser davantage
%k shape, shapelets , ye2009time
%\thi{à reformuler pour faire moins stupide}
A time series is a set of values indexed by time \cite{trirat2024universal}.
This type of data is prevalent in biology and medicine for recording physiological time series \cite{wang2016research,bar2012studying,germain2023unsupervised}, as well as in the broader context of the Internet of Things sensors \cite{fathy2018large} and measurements in physical systems \cite{luo2021deep}.
 However, due to the nature of such data, which have different sample sizes and incorporate a temporal evolution, classical statistical analysis, and machine learning methods need refinement \cite{anderson2011statistical}.
 One option to tackle this issue is to derive a feature representation of time series, which depends on the problem at hand \cite{trirat2024universal}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{"./pictures/diffeo.jpeg"}
  \vspace{-2.5em}
  \caption{A time series' graph $\msg=\{(t,s(t)): \eqsp t\in\msi\} $ can lose its structure after applying a general diffeomorphism $\phi.\msg$: a time value can be related to two values on the space axis.}
  \label{fig:diffeo}
  \vspace{-1em}
\end{figure}

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{"./pictures/transport.jpeg"}
  \vspace{-2.5em}
  \caption{LDDMM and TS-LDDMM are applied to ECG data.
  We observe that LDDMM, using a general Gaussian kernel, does not learn the time translation of the first spike but changes the space values, i.e., one spike disappears before emerging at a translated position. At the same time, TS-LDDMM handles the time change in the shape.
  This difference of \textit{deformations} implies differences in features \textit{representations}.   }
  \label{fig:transport}
  \vspace{-1em}
\end{figure*}

\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$,
by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$,
 by $||u||_\infty=\inf_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$,
and by $\Nset_{>0}$ is the set of positive integers. 
%We are in a sub class of curve quotiented by temporal reparametrisation since a time series' graph can be understood as the equivalent class of t ->(t,s(t))

%Reproducing Kernel Hilbert Space (RKHS)
\section{Methodology}
%\paragraph{Assumptions}
%Denoting by $C^0(\msi,\msj) $ the space of continous functions between two sets $\msi,\msj$,

@@

First, we introduce the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework.
 Then, we explain how to learn a discretization of the graph $\msg(\bfs_0)$ and the diffeomorphisms $(\phi_j)_{j\in[N]} $ by using LDDMM and a gradient descent minimization.
  Finally, we tackle the specificity of graph time series by deriving a representation theorem on diffeomorphisms, which enables us to select the kernel needed in LDDMM, and thus, we propose TS-LDDMM. 
%As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s_j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
 %In \Cref{section:optimization} the choice of the common time series $\mathbf{s}_0$ will be optimized to have $\phi_j$ of minimal norm. %T is given in the following part.

@@

%In all the following, we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
%\sam{Transformer , en : dans les ensembles}
\subsection{Learning diffeomorphism with LDDMM} %Thèse de Nicolas Charon
\label{section:LDDMM}
% \sam{Introduire flow normalisant et continuous normalisant, faire des exps ?}
To learn the diffeomorphisms $(\phi_j)_{j\in[N]}$, we leverage LDDMM, initially introduced in \cite{beg2005computing}.
  In a nutshell, for any $j\in [N]$, $\phi_j$ corresponds to a differential flow related to a learnable velocity field belonging to a well-chosen Reproducing Kernel Hilbert Space (RKHS).

@@

 In our practical context, $\mathbf{y}$ represents one of the \textit{discretized graphs} of the time series $(s^j)_{j\in[N]}$, defined as, for any $j\in[N]$,
 \begin{equation}
  \label{eq:descretized_graph}
  \msg(\tsig^j)=(t_i^j,\tsig^j_i)_{ i\in[n_j]}\in (\Rset^{d+1})^{n_j}
 \end{equation}
  and $\mathbf{x}$ corresponds to a discretization of the reference graph $\msg(\mathbf{s}_0)$, i.e., 
  \begin{equation}
    \label{eq:descretized_referenced_graph}
    \mathbf{x}=\tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} \eqsp . 
  \end{equation} 
% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{align}
  \label{eq:LDDMM_dynamic}
    &\frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,\\
    &\phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{align}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

@@

        \label{section:time_series_specificity}
        In this section, we tailor the LDDMM framework to handle time series data.
          The reason is that applying a general diffeomorphism $\phi$ from $\Rset^{d+1}$ to a time series' graph $\msg(s)$ can result in a set $\phi.\msg(s)$ that does not correspond to the graph of any time series, as illustrated in the \Cref{fig:diffeo}.

@@

More precisely, given two sets $\msg_0=(g_i^0)_{i\in[T_0]},\msg_1=(g_i^1)_{i\in[T_1]}\in (\Rset^{d+1})^{T_1}$ and a kernel\footnote{$\mathbb{S}^d=\{x\in\Rset^{d+1}:\eqsp |x|=1\}$} $k:(\Rset^{d+1} \times \mathbb{S}^d)^2\to \Rset$
verifying \citep[Proposition 2 \& 4]{kaltenmark2017general}, for any $\xi\in\{0,1\}$ and $i\in[T_\xi-1]$, denoting the center and length of the $i^{th}$ segment $[g_i^\xi,g_{i+1}^\xi]$ by
$c_i^\xi = (g_i^\xi + g_{i+1}^\xi)/2$, $l_i^\xi = \| g_{i+1}^\xi-g_{i}^\xi\|$, and 
$\overrightarrow{v_i}^\xi = (g_{i+1}^\xi-g_{i}^\xi)/l_i^\xi$, the varifold distance between $\msg_0$ and $\msg_1$  is defined as,
\begin{align}
  d_{\msw^*}^2(\msg_0,\msg_1) &= \sum_{i,j = 1}^{T_0-1}l^0_i k((c^0_i,\overrightarrow{v_i}^0),(c^0_j,\overrightarrow{v_j}^0))l^0_j \\
  &- 2 \sum_{i=1}^{T_0-1}\sum_{j=1}^{T_1-1}l^0_i k((c^0_i,\overrightarrow{v_i}^0),(c^1_j,\overrightarrow{v_j}^1))l^1_j \\
  &+ \sum_{i,j = 1}^{T_1-1}l^1_i k((c^1_i,\overrightarrow{v_i}^1),(c^1_j,\overrightarrow{v_j}^1))l^1_j 
\end{align}

@@

\section{Societal impact}
This paper presents work whose goal is to advance the field of Machine Learning. There are many potential societal consequences of our work, none which we feel must be specifically highlighted here.
% \section{Acknowledgements}

@@

\bibliography{ref}
\bibliographystyle{icml2023}

>>>>

@@

%Idée de parler d'inter individualité, spécialiser davantage
%k shape, shapelets , ye2009time
%\thi{à reformuler pour faire moins stupide}

\sam{Aller droit au but, se servir de la section related works pour détailler, notamment enlever le premier paragraphe dire directement que l'on veut faire du unsupervised represnetation learning for time series mais en s'intéressant par ticulièrement à la notion de shape.
Parler du contexte biomédical, définir la shape. Parler de shape analysis, introduire notre problème de représentation, dire que nous on propose une structure adapté aux time series.}
A time series is a set of values indexed by time \cite{trirat2024universal}.
This type of data is prevalent in biology and medicine for recording physiological time series \cite{wang2016research,bar2012studying,germain2023unsupervised}, as well as in the broader context of the Internet of Things sensors \cite{fathy2018large} and measurements in physical systems \cite{luo2021deep}.
 However, due to the nature of such data, which have different sample sizes and incorporate a temporal evolution, classical statistical analysis, and machine learning methods need refinement \cite{anderson2011statistical}.
 One option to tackle this issue is to derive a feature representation of time series, which depends on the problem at hand \cite{trirat2024universal}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.7\linewidth]{"./pictures/diffeo.jpeg"}

  \caption{A time series' graph $\msg=\{(t,s(t)): \eqsp t\in\msi\} $ can lose its structure after applying a general diffeomorphism $\phi.\msg$: a time value can be related to two values on the space axis.}
  \label{fig:diffeo}

\end{figure}

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{"./pictures/transport.jpeg"}

  \caption{LDDMM and TS-LDDMM are applied to ECG data.
  We observe that LDDMM, using a general Gaussian kernel, does not learn the time translation of the first spike but changes the space values, i.e., one spike disappears before emerging at a translated position. At the same time, TS-LDDMM handles the time change in the shape.
  This difference of \textit{deformations} implies differences in features \textit{representations}.   }
  \label{fig:transport}

\end{figure*}

\section{Related Works}
\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$,
by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$,
 by $||u||_\infty=\inf_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$,
and by $\Nset_{>0}$ is the set of positive integers. 
%We are in a sub class of curve quotiented by temporal reparametrisation since a time series' graph can be understood as the equivalent class of t ->(t,s(t))

%Reproducing Kernel Hilbert Space (RKHS)
\section{Background on LDDMM}
\label{section:LDDMM}
In this part, there is no novelty, we simply expose how to
 learn the diffeomorphisms $(\phi_j)_{j\in[N]}$ using LDDMM, initially introduced in \cite{beg2005computing}.
  In a nutshell, for any $j\in [N]$, $\phi_j$ corresponds to a differential flow related to a learnable velocity field belonging to a well-chosen Reproducing Kernel Hilbert Space (RKHS).

The basic problem that we consider in this section is the following. Given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$\footnote{Note that we denote by $d'\in\nset$ the ambient space}, a set of starting points $\mathbf{x}=(x_{i})_{i\in[T_1]}$ in $\Rset^{d'}$, we aim to find a diffeomorphism $\phi$ such that the finite set of points $\mathbf{y}$ is similar in a certain sense to the set of finite sets of transformed points $\phi \cdot \mathbf{x} =(\phi(x_i))_{i\in[T_1]} $.
 The function $\phi$ is occasionally referred to as a \textit{deformation}. In general, these sets $\mathbf{x},\mathbf{y}$ are meshes of continous objects, e.g surfaces, curves, images and so on.

% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

\paragraph{Discrete parametrization of diffeomorpshim.}
%\alain{adapt this without mentioning applicaiton to time seriess}
%\sam{Est ce qu'on devrait pas introduire la dicretization du graphe ici et dire que l'on peut utiliser le théoreme de représentation}
In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). % $K(x,y)\triangleq \exp(-|x-y|^2/\sigma^2)\operatorname{Id}_{d'}$ for any $(x,y)\in(\Rset^{d'})^2$, where $\sigma^2>0$ is the kernel width and $\operatorname{Id}_{d'}$ is the identity matrix.
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

\section{Methodology}
%\paragraph{Assumptions}
%Denoting by $C^0(\msi,\msj) $ the space of continous functions between two sets $\msi,\msj$,

@@

First, we introduce the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework.
 Then, we explain how to learn a discretization of the graph $\msg(\bfs_0)$ and the diffeomorphisms $(\phi_j)_{j\in[N]} $ by using LDDMM and a gradient descent minimization.
  Finally, we tackle the specificity of graph time series by deriving a representation theorem on diffeomorphisms, which enables us to select the kernel needed in LDDMM, and thus, we propose TS-LDDMM. 
  \sam{smooth this part}
  In our practical context, $\mathbf{y}$ represents one of the \textit{discretized graphs} of the time series $(s^j)_{j\in[N]}$, defined as, for any $j\in[N]$,
  \begin{equation}
   \label{eq:descretized_graph}
   \msg(\tsig^j)=(t_i^j,\tsig^j_i)_{ i\in[n_j]}\in (\Rset^{d+1})^{n_j}
  \end{equation}
   and $\mathbf{x}$ corresponds to a discretization of the reference graph $\msg(\mathbf{s}_0)$, i.e., 
   \begin{equation}
     \label{eq:descretized_referenced_graph}
     \mathbf{x}=\tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} \eqsp . 
   \end{equation}
     %\sam{Parler de l'unicté?}
     \paragraph{Coming back to the representation problem \eqref{eq:transformation}.}
     Now, we detail how to learn $(\phi^j)_{j\in[N]}$ and a discretization of the reference graph $\msg(\mathbf{s}_0)$.
      Setting $\mathbf{y}_j=\msg(\tsig^j)$ for any $j\in[N]$ and $\tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} $ as given in \eqref{eq:descretized_graph}-\eqref{eq:descretized_referenced_graph} with $\mathbf{n}_0=\operatorname{median}((n_j)_{j\in[N]})$, the representation problem given in \eqref{eq:transformation} boils down solving:
      \begin{equation}
        \label{eq:general_optimization_problem}
        \argmin_{\tilde{\msg}_0,(\alpha_k^j)_{k\in [\mathbf{n}_0]}^{j\in[N]}} \sum_{j=1}^N \mathcal{F}_{\tilde{\msg}_0,\mathbf{y}_j}\left((\alpha_k^j)_{k\in [\mathbf{n}_0]}\right)\eqsp ,
      \end{equation}
      which is carried out by a gradient descent on the control points $\tilde{\msg}_0$ and the momentums $\mathbf{\alpha}_j=(\alpha_k^j)_{k\in [\mathbf{n}_0]}$ for any $j\in[N]$, initialized by a dataset's time series graph of size $\mathbf{n}_0$ and by $0_{(d+1)\mathbf{n}_0}$ respectively.
      The optimization hyperparameter details are given in \Cref{appendix:optimizers_details}.
     The result of the minimization $\tilde{\msg}_0$ is then considered as the $\mathbf{n}_0$-samples of a common time series $\mathbf{s}_0$ and the momentums $\mathbf{\alpha}_j$ encoding $\phi_j$ yields a feature vector in $\Rset^{d \mathbf{n}_0} $ of $s^j$ for any $j\in[N]$.
     %, which will be the representation of the time series when choosing the time series' graph sets as inputs, is dependent on the choice of kernel $K$ and the set distance $L$.
     Finally, the vectors $(\mathbf{\alpha}_j)_{j\in[N]}$ can be analyzed with any statistical or machine learning tools such as Principal Components Analysis (PCA), Latent Discriminant Analysis (LDA), longitudinal data analysis and so on.
%As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s_j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
 %In \Cref{section:optimization} the choice of the common time series $\mathbf{s}_0$ will be optimized to have $\phi_j$ of minimal norm. %T is given in the following part.

@@

%In all the following, we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
%\sam{Transformer , en : dans les ensembles}
\subsection{Learning diffeomorphism with LDDMM} %Thèse de Nicolas Charon
\label{section:LDDMM}
\sam{In this part, there is nearly no novelty, we simply expose how} to
 learn the diffeomorphisms $(\phi_j)_{j\in[N]}$ using LDDMM, initially introduced in \cite{beg2005computing}.
  In a nutshell, for any $j\in [N]$, $\phi_j$ corresponds to a differential flow related to a learnable velocity field belonging to a well-chosen Reproducing Kernel Hilbert Space (RKHS).

@@

 In our practical context, $\mathbf{y}$ represents one of the \textit{discretized graphs} of the time series $(s^j)_{j\in[N]}$, defined as, for any $j\in[N]$,
 \begin{equation}
  \label{eq:descretized_graph}
  \msg(\tsig^j)=(t_i^j,\tsig^j_i)_{ i\in[n_j]}\in (\Rset^{d+1})^{n_j}
 \end{equation}
  and $\mathbf{x}$ corresponds to a discretization of the reference graph $\msg(\mathbf{s}_0)$, i.e., 
  \begin{equation}
    \label{eq:descretized_referenced_graph}
    \mathbf{x}=\tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} \eqsp . 
  \end{equation} 
% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

@@

        \label{section:time_series_specificity}
        \sam{In this section, we present our theoretical contribution: we tailor the LDDMM framework to handle time series data}.
          The reason is that applying a general diffeomorphism $\phi$ from $\Rset^{d+1}$ to a time series' graph $\msg(s)$ can result in a set $\phi.\msg(s)$ that does not correspond to the graph of any time series, as illustrated in the \Cref{fig:diffeo}.

@@

More precisely, given two sets $\msg_0=(g_i^0)_{i\in[T_0]},\msg_1=(g_i^1)_{i\in[T_1]}\in (\Rset^{d+1})^{T_1}$ and a kernel\footnote{$\mathbb{S}^d=\{x\in\Rset^{d+1}:\eqsp |x|=1\}$} $k:(\Rset^{d+1} \times \mathbb{S}^d)^2\to \Rset$
verifying \citep[Proposition 2 \& 4]{kaltenmark2017general}, for any $\xi\in\{0,1\}$ and $i\in[T_\xi-1]$, denoting the center and length of the $i^{th}$ segment $[g_i^\xi,g_{i+1}^\xi]$ by
$c_i^\xi = (g_i^\xi + g_{i+1}^\xi)/2$, $l_i^\xi = \| g_{i+1}^\xi-g_{i}^\xi\|$, and 
$\overrightarrow{v_i}^\xi = (g_{i+1}^\xi-g_{i}^\xi)/l_i^\xi$, the varifold distance between $\msg_0$ and $\msg_1$  is defined as,
\begin{align}
  &d_{\msw^*}^2(\msg_0,\msg_1) = \sum_{i,j = 1}^{T_0-1}l^0_i k((c^0_i,\overrightarrow{v_i}^0),(c^0_j,\overrightarrow{v_j}^0))l^0_j
  - 2 \sum_{i=1}^{T_0-1}\sum_{j=1}^{T_1-1}l^0_i k((c^0_i,\overrightarrow{v_i}^0),(c^1_j,\overrightarrow{v_j}^1))l^1_j \\
  &+ \sum_{i,j = 1}^{T_1-1}l^1_i k((c^1_i,\overrightarrow{v_i}^1),(c^1_j,\overrightarrow{v_j}^1))l^1_j 
\end{align}

@@

%\section{Societal impact}
%This paper presents work whose goal is to advance the field of Machine Learning. There are many potential societal consequences of our work, none which we feel must be specifically highlighted here.
% \section{Acknowledgements}

@@
